---
title: "Sampling"
---

Follow along with the demo to learn how to sample data from a population using R and the Tidyverse. You should get started by loading the tidyverse in a new R block.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```


### Task 2: Load 1,000 rows of data with the ages of people from 18-80

```{r}
set.seed(38402)

population <- tibble(
  age = sample(18:80, 1000, replace = TRUE)
)
```



### Task 3: Sample 100 rows of data from the dataset

```{r}
sample50 <- population |> 
  sample_n(100)

```



### Task 4: Calculate the mean age and standard deviation of the sample

```{r}
sample50 |> 
  summarize(
    mean_age = mean(age),
    sd_age = sd(age)
  )


```



### Task 5: Calculate the mean age and standard deviation of the population

```{r}

population |> 
  summarize(
    mean_age = mean(age),
    sd_age = sd(age)
  )

```



### Task 6: Compare the sample and population statistics

Compare the means and standard deviations. Are they similar? Go back and re-run the sampling and statistics calculations with a smaller sample size. How do the results change?

The means and standard deviations are very close together. The difference in the mean is only .27 and the difference in the standard deviation is only about .05. Theyn are very similar and the sample represents the population. 

I did it a second time with a sample of 50. This time, the difference in age is about 2.03. The difference in standard deviation is about 1.52. This shows how smaller samples can be less representative of a full population.




When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
