---
title: "HW3 WH Salaries"
---

In this assignment, we will be working with a dataset containing information about the salaries of White House staffers as of July 2026.

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2026.csv. You will create a dataframe called `wh_salaries` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
wh_salaries <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2026.csv")
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, what stands out to you about this data? Are there any salary values that will cause problems for analysis?

The first column of the data shows the first name, middle initial, and last name of the employees. It then shows whether they're a detailee or full employee of the White House, and then the salary per year as the job position of the individual.

I was initially surprised at how many five figure salaries there were, especially those between $55,000-70,000. Working in a prestigious building like the White House for the U.S. government  struck me as something that would pay more even at entry levels.

There are a few salaries listed at $0 that will provide a little bit of a skew to the calculations.



ANSWER HERE

### Task 4: Calculate the Mean and Median of the Salaries

Write code to calculate the mean and median of the salaries in the dataset using the `summarize()` function. You don't need to make a new dataframe, just display the results.

```{r}
wh_salaries |> summarize(mean_salary = mean(SALARY), median_salary = median(SALARY))
```

### Task 5: Add in the minimum and maximum salaries

Take the code you just wrote and add in the minimum and maximum salaries (the functions for that are `min()` and `max()`). Your results should show four columns: mean_salary, median_salary, min_salary, max_salary.

```{r}
wh_salaries |> summarize(mean_salary = mean(SALARY), median_salary = median(SALARY), min_salary = min(SALARY), max_salary = max(SALARY))
```

### Task 6: Describe the Results

Write a couple of sentences describing the results of your analysis. What do the mean and median salaries tell you about the data? What do the minimum and maximum salaries tell you? Are there any outliers in the data?

The mean salary is $112,703 and the median salary is $109,243. The outliers are those with a salary of 0. 

The mean being slightly higher then the median shows that there is a cluster of higher salaries for higher ranking officials that brings the mean up. 9 people with the maximum salary of $197,200 shows that there's likely a pay ceiling at that number for White House employees The 0 minimum shows that some people do not take salaries for their specific position.

### Task 7: Story Idea

Based on the data, what is your best story idea? What is the next step you would take to develop that story idea?

My story idea would be to see whether similar positions in different fields actually may offer higher salaries compared to the White House. While certain positions won't have any overlap to the employment world, many of these listed positions will.

I would do online research to find salaries of similar positions in other work fields 
